---
output: github_document
header-includes:
   - \usepackage{amsmath}
   - \usepackage{amssymb}
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = NULL
)
library(tabula)
```

# tabula <img width=120px src="man/figures/logo.svg" align="right" />

[![Build Status](https://travis-ci.org/nfrerebeau/tabula.svg?branch=master)](https://travis-ci.org/nfrerebeau/tabula)
[![codecov](https://codecov.io/gh/nfrerebeau/tabula/branch/master/graph/badge.svg)](https://codecov.io/gh/nfrerebeau/tabula)
[![GitHub Release](https://img.shields.io/github/release/nfrerebeau/tabula.svg)](https://github.com/nfrerebeau/tabula/releases)
[![CRAN Version](http://www.r-pkg.org/badges/version/tabula)](https://cran.r-project.org/package=tabula)
[![CRAN Downloads](http://cranlogs.r-pkg.org/badges/tabula)](https://cran.r-project.org/package=tabula)
[![lifecycle](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.1489944.svg)](https://doi.org/10.5281/zenodo.1489944)

## Overview

`tabula` provides an easy way to examine archaeological count data (artifacts, faunal remains, etc.). This package includes several measures of diversity: e.g. richness and rarefaction (Chao1, Chao2, ACE, ICE, etc.), diversity/dominance and evenness (Brillouin, Shannon, Simpson, etc.), turnover and similarity (Brainerd-Robinson, ...). It also provides matrix seriation methods (reciprocal ranking, CA-based seriation) for chronological modeling and dating. The package make it easy to visualize count data and statistical thresholds: rank/abundance plots, Ford and Bertin diagrams, etc.

## Installation

You can install the released version of `tabula` from [CRAN](https://CRAN.R-project.org) with:

```{r cran-installation, eval = FALSE}
install.packages("tabula")
```

Or install the development version from GitHub with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("nfrerebeau/tabula")
```

## Usage

`tabula` provides a set of S4 classes that extend the `matrix` data type from R `base`. These new classes represent different special types of matrix.

* Abundance matrix:
    * `CountMatrix` represents count data,
    * `FrequencyMatrix` represents relative frequency data.
* Logical matrix:
    * `IncidenceMatrix` represents presence/absence data.
* Other numeric matrix:
    * `OccurrenceMatrix` represents a co-occurence matrix.
    * `SimilarityMatrix` represents a (dis)similarity matrix.

It assumes that you keep your data tidy: each variable (type/taxa) must be saved in its own column and each observation (sample/case) must be saved in its own row.

These new classes are of simple use, on the same way as the base `matrix`:

```{r create, eval=FALSE}
# Define a count data matrix
quanti <- CountMatrix(data = sample(0:10, 100, TRUE),
                      nrow = 10, ncol = 10)

# Define a logical matrix
# Data will be coerced with as.logical()
quali <- IncidenceMatrix(data = sample(0:1, 100, TRUE),
                         nrow = 10, ncol = 10)
```

`tabula` uses coercing mechanisms (with validation methods) for data type conversions:

```{r coerce}
# Create a count matrix
A1 <- CountMatrix(data = sample(0:10, 100, TRUE),
                  nrow = 10, ncol = 10)

# Coerce counts to frequencies
B <- as(A1, "FrequencyMatrix")

# Row sums are internally stored before coercing to a frequency matrix
# (use totals() to get these values)
# This allows to restore the source data
A2 <- as(B, "CountMatrix")
all(A1 == A2)

# Coerce to presence/absence
C <- as(A1, "IncidenceMatrix")

# Coerce to a co-occurrence matrix
D <- as(A1, "OccurrenceMatrix")
```

Several types of graphs are available in `tabula` which uses `ggplot2` for plotting informations. This makes it easy to customize diagramms (e.g. using themes and scales).

Spot matrix (easier to read than a heatmap [^1]) allows direct examination of data (above/below some threshold):

```{r plot-freq, fig.width=4, fig.height=6, fig.align="center"}
# Plot frequencies with the column means as a threshold
ceram_counts <- as(mississippi, "CountMatrix")
plotSpot(ceram_counts, threshold = mean) +
  ggplot2::labs(size = "Frequency", colour = "Mean") +
  khroma::scale_colour_vibrant()
```

```{r plot-occ, fig.width=6, fig.height=4, fig.align="center"}
# Plot co-occurence of types
# (i.e. how many times (percent) each pairs of taxa occur together 
# in at least one sample.)
ceram_occ <- as(mississippi, "OccurrenceMatrix")
plotSpot(ceram_occ) +
  ggplot2::labs(size = "", colour = "Co-occurrence") +
  ggplot2::theme(legend.box = "horizontal") +
  khroma::scale_colour_YlOrBr()
```

Bertin or Ford (battleship curve) diagramms can be plotted, with statistic threshold (B. Desachy's [*sériographe*](https://doi.org/10.3406/pica.2004.2396)). The positive difference from the column mean percentage (in french "écart positif au pourcentage moyen", EPPM) represents a deviation from the situation of statistical independence. EPPM is a usefull graphical tool to explore significance of relationship between rows and columns related to seriation.

```{r seriograph, fig.width=7, fig.height=3.5, fig.align="center"}
counts <- as(compiegne, "CountMatrix")
plotBar(counts, EPPM = TRUE) +
  khroma::scale_fill_bright()
```

### Seriation

```{r seriate-incidence}
# Build an incidence matrix with random data
set.seed(12345)
incidence <- IncidenceMatrix(data = sample(0:1, 400, TRUE, c(0.6, 0.4)),
                             nrow = 20)

# Get seriation order on rows and columns
# Correspondance analysis-based seriation
(indices <- seriate(incidence, method = "correspondance", margin = c(1, 2)))

# Permute matrix rows and columns
incidence2 <- permute(incidence, indices)
```

```{r permute-incidence, fig.width=3.5, fig.height=3.5, fig.show="hold"}
# Plot matrix
plotMatrix(incidence) + 
  ggplot2::labs(title = "Original matrix") +
  ggplot2::scale_fill_manual(values = c("TRUE" = "black", "FALSE" = "white"))
plotMatrix(incidence2) + 
  ggplot2::labs(title = "Rearranged matrix") +
  ggplot2::scale_fill_manual(values = c("TRUE" = "black", "FALSE" = "white"))
```

### Dating

This package provides an implementation of the chronological modeling method developed by Bellanger and Husi ([2012](https://doi.org/10.1016/j.jas.2011.06.031)). This allows the construction of two different probability estimate
density curves of archaeological assembalge dates (the so-called *event* and *accumulation* dates). Note that this implementation is experimental (see `help(dateEvent)`).

```{r date, fig.width=3.5, fig.height=3.5, fig.show="hold"}
# Coerce dataset to abundance (count) matrix
zuni <- as(zuni, "CountMatrix")
# Assume that some assemblages are reliably dated (this is NOT a real example).
# The names of the vector entries must match the names of the assemblages.
setDates(zuni) <- c(
  LZ0569 = 1097, LZ0279 = 1119, CS16 = 1328, LZ0066 = 1111,
  LZ0852 = 1216, LZ1209 = 1251, CS144 = 1262, LZ0563 = 1206,
  LZ0329 = 1076, LZ0005Q = 859, LZ0322 = 1109, LZ0067 = 863,
  LZ0578 = 1180, LZ0227 = 1104, LZ0610 = 1074
)
# Model the event and accumulation date for each assemblage.
model <- dateEvent(zuni, cutoff = 90)
# Plot event and accumulation date distributions
plotDate(model, select = "LZ1105") +
  ggplot2::labs(title = "Activity plot") +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom")
plotDate(model, select = "LZ1105", tempo = TRUE) +
  ggplot2::labs(title = "Tempo plot") +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom")
```

### Analysis

*Diversity* can be measured according to several indices (sometimes refered to as indices of *heterogeneity*):

```{r diversity}
H <- diversity(ceram_counts, method = c("shannon", "brillouin", "simpson", 
                                        "mcintosh", "berger"), simplify = TRUE)
head(H)
```

Note that `berger`, `mcintosh` and `simpson` methods return a *dominance* index, not the reciprocal form usually adopted, so that an increase in the value of the index accompanies a decrease in diversity.

Corresponding *evenness* (i.e. a measure of how evenly individuals are distributed across the sample) can also be computed.

Several methods can be used to acertain the degree of *turnover* in taxa composition along a gradient on qualitative (presence/absence) data. It assumes that the order of the matrix rows (from $1$ to $n$) follows the progression along the gradient/transect.

Diversity can also be measured by addressing *similarity* between pairs of sites:

```{r similarity-brainerd, fig.width=6, fig.height=5, fig.align="center"}
# Brainerd-Robinson index
S <- similarity(ceram_counts, method = "brainerd")

# Plot the similarity matrix
plotSpot(S) +
  ggplot2::labs(size = "Similarity", colour = "Similarity") +
  khroma::scale_colour_YlOrBr()
```

Ranks *vs* abundance plot can be used for abundance models (model fitting will be implemented in a futur release):

```{r plot-rank, fig.width=7, fig.height=3.5, fig.align="center"}
plotRank(counts, log = "xy") +
  ggplot2::theme_bw()
```

## Contributing
Please note that the `tabula` project is released with a [Contributor Code of Conduct](CODE_OF_CONDUCT.md). By contributing to this project, you agree to abide by its terms.

[^1]: Adapted from Dan Gopstein's original [spot matrix](https://dgopstein.github.io/articles/spot-matrix/).
