---
output: github_document
header-includes:
   - \usepackage{amsmath}
   - \usepackage{amssymb}
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
library(tabula)
```

# tabula <img width=120px src="man/figures/logo.svg" align="right" />

[![Build Status](https://travis-ci.org/nfrerebeau/tabula.svg?branch=master)](https://travis-ci.org/nfrerebeau/tabula)
[![codecov](https://codecov.io/gh/nfrerebeau/tabula/branch/master/graph/badge.svg)](https://codecov.io/gh/nfrerebeau/tabula)
[![GitHub Release](https://img.shields.io/github/release/nfrerebeau/tabula.svg)](https://github.com/nfrerebeau/tabula/releases)
[![CRAN Version](http://www.r-pkg.org/badges/version/tabula)](https://cran.r-project.org/package=tabula)
[![CRAN Downloads](http://cranlogs.r-pkg.org/badges/tabula)](https://cran.r-project.org/package=tabula)
[![lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.1489944.svg)](https://doi.org/10.5281/zenodo.1489944)

## Overview

`tabula` provides an easy way to examine archaeological count data (artifacts, faunal remains, etc.). This package includes several measures of diversity: e.g. richness and rarefaction (Chao1, Chao2, ACE, ICE, etc.), diversity/dominance and evenness (Brillouin, Shannon, Simpson, etc.), turnover and similarity (Brainerd-Robinson, ...). It also provides matrix seriation methods (reciprocal ranking, CA-based seriation) for chronological modeling and dating. The package make it easy to visualize count data and statistical thresholds: rank/abundance plots, Ford and Bertin diagrams, etc.

## Installation

You can install the released version of `tabula` from [CRAN](https://CRAN.R-project.org) with:

```{r cran-installation, eval = FALSE}
install.packages("tabula")
```

Or install the development version from GitHub with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("nfrerebeau/tabula")
```

## Usage

`tabula` provides a set of S4 classes that extend the `matrix` data type from R `base`. These new classes represent different special types of matrix.

* Abundance matrix:
    * `CountMatrix` represents count data,
    * `FrequencyMatrix` represents relative frequency data.
* Logical matrix:
    * `IncidenceMatrix` represents presence/absence data.
    * `OccurrenceMatrix` represents a co-occurence matrix.
* Numeric matrix:
    * `SimilarityMatrix` represents a (dis)similarity matrix.

It assumes that you keep your data tidy: each variable (type/taxa) must be saved in its own column and each observation (sample/case) must be saved in its own row.

These new classes are of simple use, on the same way as the base `matrix`:

```{r create, eval=FALSE}
# Define a count data matrix
quanti <- CountMatrix(data = sample(0:10, 100, TRUE),
                      nrow = 10, ncol = 10)

# Define a logical matrix
# Data will be coerced with as.logical()
quali <- IncidenceMatrix(data = sample(0:1, 100, TRUE),
                         nrow = 10, ncol = 10)
```

`tabula` uses coercing mechanisms (with validation methods) for data type conversions:

```{r coerce}
# Create a count matrix
A1 <- CountMatrix(data = sample(0:10, 100, TRUE),
                  nrow = 10, ncol = 10)

# Coerce counts to frequencies
B <- as(A1, "FrequencyMatrix")

# Row sums are internally stored before coercing to a frequency matrix
# (use totals() to get these values)
# This allows to restore the source data
A2 <- as(B, "CountMatrix")
all(A1 == A2)

# Coerce to presence/absence
C <- as(A1, "IncidenceMatrix")

# Coerce to a co-occurrence matrix
D <- as(A1, "OccurrenceMatrix")
```

### Analysis
#### Sample richness

```{r richness}
compiegne <- as(compiegne, "CountMatrix")
richness(compiegne, method = "chao1")
```

#### Heterogeneity and evenness measures

*Diversity* can be measured according to several indices (sometimes refered to as indices of *heterogeneity*):

```{r diversity}
diversity(compiegne, method = c("shannon", "brillouin", "simpson", "mcintosh", "berger"), simplify = TRUE)
```

Note that `berger`, `mcintosh` and `simpson` methods return a *dominance* index, not the reciprocal form usually adopted, so that an increase in the value of the index accompanies a decrease in diversity.

Corresponding *evenness* (i.e. a measure of how evenly individuals are distributed across the sample) can also be computed.

#### Turnover and Similarity

The several methods can be used to acertain the degree of *turnover* in taxa composition along a gradient ($\beta$-diversity) on qualitative (presence/absence) data. It assumes that the order of the matrix rows (from $1$ to $n$) follows the progression along the gradient/transect.

$\beta$-diversity can also be measured by addressing *similarity* between pairs of sites:

```{r similarity-brainerd}
similarity(compiegne, method = "brainerd")
```

### Seriation

```{r seriate-incidence}
# Build an incidence matrix with random data
incidence <- IncidenceMatrix(data = sample(0:1, 400, TRUE, c(0.6, 0.4)),
                             nrow = 20)

# Get seriation order on rows and columns
# Correspondance analysis-based seriation
set.seed(12345)
(indices <- seriate(incidence, method = "correspondance", margin = c(1, 2)))

# Permute matrix rows and columns
incidence2 <- permute(incidence, indices)
```

```{r permute-incidence, eval=FALSE, warning=FALSE}
# Plot matrix
library(ggplot2)
plotMatrix(incidence) + 
  labs(title = "Original matrix") +
  scale_fill_manual(values = c("TRUE" = "black", "FALSE" = "white"))
plotMatrix(incidence2) + 
  labs(title = "Rearranged matrix") +
  scale_fill_manual(values = c("TRUE" = "black", "FALSE" = "white"))
```

```{r permute-incidence-plots, echo=FALSE, warning=FALSE, fig.height=3.5}
# Plot matrix
cowplot::plot_grid(
  plotMatrix(incidence) + 
    ggplot2::labs(title = "Original matrix") +
    ggplot2::scale_fill_manual(values = c("TRUE" = "black", "FALSE" = "white")),
  plotMatrix(incidence2) + 
    ggplot2::labs(title = "Rearranged matrix") +
    ggplot2::scale_fill_manual(values = c("TRUE" = "black", "FALSE" = "white")),
  ncol = 2
)
```

### Visualization

`tabula` makes an extensive use of `ggplot2` for plotting informations. This makes it easy to customize diagramms (e.g. using themes and scales).

Bertin of Ford (battleship curve) diagramms can be plotted, with statistic threshold ([B. Desachy's *sériographe*](https://doi.org/10.3406/pica.2004.2396)). The positive difference from the column mean percentage (in french "écart positif au pourcentage moyen", EPPM) represents a deviation from the situation of statistical independence. EPPM is a usefull graphical tool to explore significance of relationship between rows and columns related to seriation.

```{r seriograph, fig.height=3.5}
count <- as(compiegne, "CountMatrix")
plotBar(count, EPPM = TRUE)
```


Matrix plot is displayed as a heatmap. The PVI matrix ([B. Desachy's *matrigraphe*](https://doi.org/10.3406/pica.2004.2396)) allows to explore deviations from independence (an intuitive graphical approach to $\chi^2$).

```{r matrigraph, fig.height=3.5}
plotMatrix(count, PVI = TRUE) +
  ggplot2::scale_fill_gradient2(midpoint = 1)
```

Spot matrix (easier to read than a heatmap [^1]) allows direct examination of data (above/below some threshold):

```{r spot, fig.height=3.5}
plotSpot(count, threshold = mean)
```

Ranks *vs* abundance plot can be used for abundance models (model fitting will be implemented in a futur release):

```{r rank, fig.height=3.5}
plotRank(count, log = "xy")
```

[^1]: Adapted from Dan Gopstein's original [spot matrix](https://dgopstein.github.io/articles/spot-matrix/).

## Contributing
Please note that the `tabula` project is released with a [Contributor Code of Conduct](CODE_OF_CONDUCT.md). By contributing to this project, you agree to abide by its terms.
