---
title: "Matrix classes"
author: "N. Frerebeau"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    number_sections: yes
    fig_caption: yes
    toc: true
header-includes:
   - \usepackage{amsmath}
   - \usepackage{amssymb}
bibliography: bibliography.bib
vignette: >
  %\VignetteIndexEntry{Matrix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, echo=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(tabula)
```

# Definitions
`tabula` provides a set of S4 classes that extend the `matrix` data type from R `base`. These new classes represent different special types of matrix.

* Abundance matrix:
    * `CountMatrix` represents count data,
    * `FrequencyMatrix` represents relative frequency data.
* Logical matrix:
    * `IncidenceMatrix` represents presence/absence data.
* Numeric matrix:
    * `OccurrenceMatrix` represents a co-occurence matrix.
    * `SimilarityMatrix` represents a (dis)similarity matrix.
    
It assumes that you keep your data tidy: each variable (taxon/type) must be saved in its own column and each observation (sample/case) must be saved in its own row. Missing values are not allowed.

## Abundance matix
### Count matrix

We denote the $m \times p$ count matrix by $A = \left[ a_{ij} \right] ~\forall i \in \left[ 1,m \right], j \in \left[ 1,p \right]$ with row and column sums:

\begin{align}
 a_{i \cdot} = \sum_{j = 1}^{p} a_{ij} &&
 a_{\cdot j} = \sum_{i = 1}^{m} a_{ij} &&
 a_{\cdot \cdot} = \sum_{i = 1}^{m} \sum_{j = 1}^{p} a_{ij} &&
 \forall a_{ij} \in \mathbb{N}
\end{align}

### Frequency matrix

A frequency matrix represents relative abundances.

We denote the $m \times p$ frequency matrix by $B = \left[ b_{ij} \right] ~\forall i \in \left[ 1,m \right], j \in \left[ 1,p \right]$ with row and column sums:

\begin{align}
 b_{i \cdot} = \sum_{j = 1}^{p} b_{ij} = 1 &&
 b_{\cdot j} = \sum_{i = 1}^{m} b_{ij} &&
 b_{\cdot \cdot} = \sum_{i = 1}^{m} \sum_{j = 1}^{p} b_{ij} &&
 \forall b_{ij} \in \left[ 0,1 \right]
\end{align}

## Logical matrix
### Incidence matrix

We denote the $m \times p$ incidence matrix by $C = \left[ c_{ij} \right] ~\forall i \in \left[ 1,m \right], j \in \left[ 1,p \right]$ with row and column sums:

\begin{align}
 c_{i \cdot} = \sum_{j = 1}^{p} c_{ij} &&
 c_{\cdot j} = \sum_{i = 1}^{m} c_{ij} &&
 c_{\cdot \cdot} = \sum_{i = 1}^{m} \sum_{j = 1}^{p} c_{ij} &&
 \forall c_{ij} \in \lbrace 0,1 \rbrace
\end{align}

## Numeric matrix
### Co-occurrence matrix

A co-occurrence matrix is a symetric matrix with zeros on its main diagonal, which works out how many times (expressed in percent) each pairs of taxa occur together in at least one sample.

The $p \times p$ co-occurrence matrix $D = \left[ d_{i,j} \right] ~\forall i,j \in \left[ 1,p \right]$ is defined over an $m \times p$ abundance matrix $A = \left[ a_{x,y} \right] ~\forall x \in \left[ 1,m \right], y \in \left[ 1,p \right]$ as:

$$ d_{i,j} = \sum_{x = 1}^{m} \bigcap_{y = i}^{j} a_{xy} $$

with row and column sums:

\begin{align}
  d_{i \cdot} = \sum_{j \geqslant i}^{p} d_{ij} &&
  d_{\cdot j} = \sum_{i \leqslant j}^{p} d_{ij} &&
  d_{\cdot \cdot} = \sum_{i = 1}^{p} \sum_{j \geqslant i}^{p} d_{ij} &&
  \forall d_{ij} \in \mathbb{N}
\end{align}

# Usage
## Create

These new classes are of simple use, on the same way as the base `matrix`:

```{r create}
set.seed(12345)
# Create a count data matrix
CountMatrix(data = sample(0:10, 100, TRUE),
            nrow = 10, ncol = 10)

# Create an incidence (presence/absence) matrix
# Numeric values are coerced to logical as by as.logical
IncidenceMatrix(data = sample(0:1, 100, TRUE),
                nrow = 10, ncol = 10)
```

Note that a `FrequencyMatrix` can only be created by coercion from a `CountMatrix` to ensure data integrity (see below).

## Coerce

`tabula` uses coercing mechanisms (with validation methods) for data type conversions:

```{r coerce}
# Create a count matrix
# Numeric values are coerced to integer and hence truncated towards zero
set.seed(12345)
A1 <- CountMatrix(data = sample(0:10, 100, TRUE),
                  nrow = 10, ncol = 10)

# Coerce counts to frequencies
B <- as(A1, "FrequencyMatrix")

# Row sums are internally stored before coercing to a frequency matrix
totals(B)
# This allows to restore the source data
A2 <- as(B, "CountMatrix")
all(A1 == A2)

# Coerce to presence/absence
C <- as(A1, "IncidenceMatrix")

# Coerce to a co-occurrence matrix
D <- as(A1, "OccurrenceMatrix")
```

# Visualization

Several types of graphs are available in `tabula` which uses `ggplot2` for plotting informations. This makes it easy to customize diagramms (e.g. using themes and scales).

## Spot plot
Spot matrix allows direct examination of data (above/below some threshold):

```{r plot-freq, fig.width=4, fig.height=6, fig.align="center"}
# Plot frequencies with the column means as a threshold
mississippi_counts <- as(mississippi, "CountMatrix")

plotSpot(mississippi_counts, threshold = mean) +
  ggplot2::labs(size = "Frequency", colour = "Mean") +
  khroma::scale_colour_vibrant()
```

```{r plot-occ, fig.width=6, fig.height=4, fig.align="center"}
# Plot co-occurence of types
# (i.e. how many times (percent) each pairs of taxa occur together 
# in at least one sample.)
mississippi_occ <- as(mississippi, "OccurrenceMatrix")
plotSpot(mississippi_occ) +
  ggplot2::labs(size = "", colour = "Co-occurrence") +
  ggplot2::theme(legend.box = "horizontal") +
  khroma::scale_colour_YlOrBr()
```

## Matrix plot

Abundance matrix can be displayed as a heatmap of relative abundances (frequency), or as percentages of the independence value (in french, "pourcentages de valeur d'indépendance", PVI).

```{r plot-matrix, fig.width=7, fig.height=3.5, fig.align="center"}
# Reproduce B. Desachy's matrigraphe
boves_counts <- as(boves, "CountMatrix")

plotMatrix(boves_counts) +
  ggplot2::theme_light() +
  khroma::scale_fill_YlOrBr()
```

PVI is calculated for each cell as the percentage to the column theoretical independence value: PVI greater than $1$ represent positive deviations from the independance, whereas PVI smaller than $1$ represent negative deviations [@desachy2004]. The PVI matrix allows to explore deviations from independence (an intuitive graphical approach to $\chi^2$), in such a way that a high-contrast matrix has quite significant deviations, with a low risk of being due to randomness [@desachy2004].

```{r plot-matrigraphe, fig.width=7, fig.height=3.5, fig.align="center"}
# Reproduce B. Desachy's matrigraphe
plotMatrix(boves_counts, PVI = TRUE) +
  ggplot2::scale_fill_gradient2(midpoint = 1) +
  ggplot2::theme_bw()
```

## Bar plot
@bertin1977 or @ford1962 (battleship curve) diagramms can also be plotted, with statistic threshold.

```{r plot-bertin, fig.width=7, fig.height=7, fig.align="center"}
plotBar(boves_counts, center = FALSE, horizontal = FALSE) +
  ggplot2::labs(title = "Ford diagram") +
  ggplot2::theme_bw()
```

```{r plot-ford, fig.width=7, fig.height=3.5, fig.align="center"}
plotBar(boves_counts, center = TRUE, horizontal = FALSE) +
  ggplot2::labs(title = "Bertin diagram") +
  ggplot2::theme_light()
```

The positive difference from the column mean percentage (in french "écart positif au pourcentage moyen", EPPM) represents a deviation from the situation of statistical independence [@desachy2004]. As independence can be interpreted as the absence of relationships between types and the chronological order of the assemblages, EPPM is a usefull graphical tool to explore significance of relationship between rows and columns related to seriation [@desachy2004].

```{r plot-seriographe, fig.width=7, fig.height=3.5, fig.align="center"}
# Reproduce B. Desachy's sériographe
plotBar(boves_counts, EPPM = TRUE) +
  khroma::scale_fill_bright()
```

# References
