---
title: "Matrix classes"
author: "N. Frerebeau"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
    toc: true
header-includes:
   - \usepackage{amsmath}
   - \usepackage{amssymb}
vignette: >
  %\VignetteIndexEntry{Matrix classes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, echo=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(tabula)
```

## Overview

`tabula` provides a set of S4 classes that extend the `matrix` data type from R `base`. These new classes represent different special types of matrix.

* Abundance matrix:
    * `CountMatrix` represents count data,
    * `FrequencyMatrix` represents frequency data.
* Logical matrix:
    * `IncidenceMatrix` represents presence/absence data.
    * `OccurrenceMatrix` represents a co-occurence matrix.
    
It assumes that you keep your data tidy: each variable (type) must be saved in its own column and each observation (case) must be saved in its own row.

Missing values are not allowed.

## Definitions
### Abundance matix
#### Count matrix

We denote the $m \times p$ count matrix by $A = \left[ a_{ij} \right] ~\forall i \in \left[ 1,m \right], j \in \left[ 1,p \right]$ with row and column sums:

\begin{align}
 a_{i \cdot} = \sum_{j = 1}^{p} a_{ij} &&
 a_{\cdot j} = \sum_{i = 1}^{m} a_{ij} &&
 a_{\cdot \cdot} = \sum_{i = 1}^{m} \sum_{j = 1}^{p} a_{ij} &&
 \forall a_{ij} \in \mathbb{N}
\end{align}

#### Frequency matrix

A frequency matrix represents relative abundances.

We denote the $m \times p$ frequency matrix by $B = \left[ b_{ij} \right] ~\forall i \in \left[ 1,m \right], j \in \left[ 1,p \right]$ with row and column sums:

\begin{align}
 b_{i \cdot} = \sum_{j = 1}^{p} b_{ij} = 1 &&
 b_{\cdot j} = \sum_{i = 1}^{m} b_{ij} &&
 b_{\cdot \cdot} = \sum_{i = 1}^{m} \sum_{j = 1}^{p} b_{ij} &&
 \forall b_{ij} \in \left[ 0,1 \right]
\end{align}

### Logical matrix
#### Incidence matrix

We denote the $m \times p$ incidence matrix by $C = \left[ c_{ij} \right] ~\forall i \in \left[ 1,m \right], j \in \left[ 1,p \right]$ with row and column sums:

\begin{align}
 c_{i \cdot} = \sum_{j = 1}^{p} c_{ij} &&
 c_{\cdot j} = \sum_{i = 1}^{m} c_{ij} &&
 c_{\cdot \cdot} = \sum_{i = 1}^{m} \sum_{j = 1}^{p} c_{ij} &&
 \forall c_{ij} \in \lbrace 0,1 \rbrace
\end{align}

#### Co-occurrence matrix

A co-occurrence matrix is a symetric matrix with zeros on its main diagonal, which works out which pairs of taxa occur together in at least one sample

<!-- The $p \times p$ co-occurrence matrix $D = \left[ d_{i,j} \right] ~\forall i,j \in \left[ 1,p \right]$ is defined over an $m \times p$ incidence matrix $C = \left[ C_{x,y} \right] ~\forall x \in \left[ 1,m \right], y \in \left[ 1,p \right]$ as: -->

<!-- \[ d_{i,j} = \begin{cases}  -->
<!--       1 & \text{if}~ m_{x,i} ~\text{and}~ m_{x,j} \\ -->
<!--       0 & \text{otherwise} -->
<!--    \end{cases} -->
<!-- \] -->

<!-- with row and column sums: -->

<!-- \begin{align} -->
<!--  d_{i \cdot} = \sum_{j \geqslant i}^{p} d_{ij} && -->
<!--  d_{\cdot j} = \sum_{i \leqslant j}^{p} d_{ij} && -->
<!--  d_{\cdot \cdot} = \sum_{i = 1}^{p} \sum_{j \geqslant i}^{p} d_{ij} && -->
<!--  \forall d_{ij} \in \lbrace 0,1 \rbrace -->
<!-- \end{align} -->

## Usage

### Create

These new classes are of simple use, on the same way as the base `matrix`:

```{r create}
# Create a count data matrix
CountMatrix(data = sample(0:10, 100, TRUE),
            nrow = 10, ncol = 10)

# Create an incidence (presence/absence) matrix
# Numeric values are coerced to logical as by as.logical
IncidenceMatrix(data = sample(0:1, 100, TRUE),
                nrow = 10, ncol = 10)
```

Note that a `FrequencyMatrix` can only be created by coercion from a `CountMatrix` to ensure data integrity (see below).

### Coerce

`tabula` uses coercing mechanisms (with validation methods) for data type conversions:

```{r coerce}
# Create a count matrix
#  Numeric values are coerced to integer and hence truncated towards zero
A1 <- CountMatrix(data = sample(0:10, 100, TRUE),
                  nrow = 10, ncol = 10)

# Coerce counts to frequencies
B <- as(A1, "FrequencyMatrix")

# Row sums are internally stored before coercing to a frequency matrix
totals(B)
# This allows to restore the source data
A2 <- as(B, "CountMatrix")
all(A1 == A2)

# Coerce to presence/absence
C <- as(A1, "IncidenceMatrix")

# Coerce to a co-occurrence matrix
D <- as(A1, "OccurrenceMatrix")
```
